---
title: "cisMR-cML for robust cis Mendelian randomization analysis"
output: rmarkdown::html_vignette
theme: unite
author: Zhaotong Lin
date: March 13, 2024
highlight: zenburn
vignette: >
  %\VignetteIndexEntry{cisMR-cML for robust cis Mendelian randomization analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---



```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette shows an example of applying cisMR-cML to infer causal relationship of LDL on CAD using cis-SNPs on the PCSK9 region.

```{r setup}
library(cisMRcML)
```

## Example 

We provide the pre-processed data for LDL and CAD on the PCSK9 region. 
```{r}
data("mr_dat")
str(mr_dat)
```
As shown above, 9 SNPs in the PCSK9 region were used in the analysis (the first real data application in the manuscript), in which 8 SNPs (shown in `mr_dat$exp_IV`) were jointly associated (selected by [GCTA-COJO](https://yanglab.westlake.edu.cn/software/gcta/#COJO)) with LDL, and 1 SNP (shown in `mr_dat$out_IV`) was associate with CAD. `mr_dat$LD_mat` is the LD/correlation matrix among the 9 SNPs, which was estimated based on the UK Biobank White-ancestry individuals.

`mr_dat$exp_df` and `mr_dat$out_df` correspond to the GWAS summary statistics for LDL (exposure) and CAD (outcome) respectively. In particular, `mr_dat$exp_df$b`, `mr_dat$exp_df$se`, `mr_dat$exp_df$p` are the raw GWAS marginal summary statistics, and `mr_dat$exp_df$bJ` is the **conditional** genetic effect estimates on the exposure (similar for `mr_dat$out_df`), which can be calculated as follows (see Manuscript Section 4.2):

```{r, eval=FALSE}
mr_dat$exp_df$cor = mr_dat$exp_df$b / sqrt(mr_dat$exp_df$b^2 + (mr_dat$exp_df$N - 2) *
                                             mr_dat$exp_df$se^2)
mr_dat$exp_df$bJ = solve(mr_dat$LD_mat) %*% mr_dat$exp_df$cor
mr_dat$out_df$cor = mr_dat$out_df$b / sqrt(mr_dat$out_df$b^2 + (mr_dat$out_df$N - 2) *
                                             mr_dat$out_df$se^2)
mr_dat$out_df$bJ = solve(mr_dat$LD_mat) %*% mr_dat$out_df$cor
```

Next, we apply cisMR-cML with 100 data perturbations with 5 random starts where $\theta^{(0)}\sim U(-0.1,0.1)$:
```{r}
b_exp_cond=mr_dat$exp_df$bJ
b_out_cond=mr_dat$out_df$bJ 
Sig_exp1 = solve(mr_dat$LD_mat) %*% (mr_dat$exp_df$se_cor %o% mr_dat$exp_df$se_cor * mr_dat$LD_mat) %*%
  solve(mr_dat$LD_mat)
Sig_out1 = solve(mr_dat$LD_mat) %*% (mr_dat$out_df$se_cor %o% mr_dat$out_df$se_cor * mr_dat$LD_mat) %*%
  solve(mr_dat$LD_mat)
Sig_exp_inv=solve(Sig_exp1)
Sig_out_inv=solve(Sig_out1)

t0 = proc.time()[3]
ciscML_res = cismr_cML_DP(b_exp=b_exp_cond,b_out=b_out_cond,
                          Sig_exp_inv=Sig_exp_inv,Sig_out_inv=Sig_out_inv,maxit=200,
                          n = mr_dat$N1,random_start = 5,
                          min_theta_range=-0.1,max_theta_range=0.1,
                          num_pert=100,random_start_pert=5,random_seed = 12345)
t1 = proc.time()[3]
cat("The running time is: ", t1 - t0, " seconds.")
```

```{r}
cat("The estimated effect of the exposure on outcome: ", ciscML_res$BIC_DP_theta)
cat("Standard error of the causal estimate: ", ciscML_res$BIC_DP_se)
cat("P-value for the causal estimate: ", ciscML_res$BIC_DP_p)
```

At this point, we have reproduced the LDL-CAD result in our manuscript.

If you failed to reproduce the results, please make sure you have the same R packages version (especially MASS_7.3-51.4) as those in the [built vignette](https://rpubs.com/ZhaotongL/cisMRcML_vignette).

```{r}
sessionInfo()
```




